### 2018-02-09 submission/manuscript.Rmd
fix study analysis overview

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased, after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in the odds ratio for individuals to have both adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma tumors (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for individuals to have carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most significantly increased odds ratios for individuals to have either adenoma or carcinoma tumors regardless of sample type. Prediction models built to classify either individuals with adenoma or carcinoma were trained on the whole community or selected genera (top 5 highest and lowest odds ratios) from either fecal or tissue samples. Both the full and select models for either adenoma or carcinoma resulted in similar classification success according to Area Under the Curve (AUC). Importantly, all models had a generalized decrease in AUC between train and test data sets. The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and tumor were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

@@

***Study Analysis Overview:***
OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and RR. The Bray-Curtis index was used to assess, for each individual study, differences between control-adenoma and control-carcinoma. Next, all common genera were assessed for differences in RR for having an adenoma or carcinoma and ranked based on P-value. We then built Random Forest models based on full or  selected community, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies. This process was repeated for every study in the meta-analysis. A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased RR based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which only control versus adenoma stage were made, the carcinoma samples were excluded from each respective study. Similarly, for comparisons in which control versus carcinoma stage were made the adenoma samples were excluded from each respective study. The data were split between feces and tissue samples. Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched). Where applicable for each study, predictions for adenoma and carcinoma stage of disease were then tested for feces, matched tissue, and unmatched tissue.

>>>>

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in the odds ratio for individuals to have both adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma tumors (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for individuals to have carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most increased odds ratios for individuals to have either adenoma or carcinoma tumors regardless of sample type. Prediction models built to classify either individuals with adenoma or carcinoma were trained on the whole community or selected genera (top 5 highest and lowest odds ratios) from either fecal or tissue samples. Both the full and select models for either adenoma or carcinoma resulted in similar classification success according to Area Under the Curve (AUC). The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and tumor were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

@@

***Study Analysis Overview:***
OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and OR. For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma. Next, all common genera were assessed for differences in OR for having an adenoma or carcinoma and ranked based on P-value. We then built Random Forest models based on the full or selected community (the top 5 increased and top 5 decreased OR based on P-value). Comparison between the full and selected models took two different approaches. In the first approach, models were trained on one study then tested on the remaining studies. This process was repeated for every study in the meta-analysis. In the second approach models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased OR based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which only control versus adenoma indivdiuals were made, the carcinoma individuals were excluded from each respective study. Similarly, for comparisons in which control versus carcinoma individuals were made the adenoma individuals were excluded from each respective study. For all analysis completed feces and tissue samples were kept seperate. Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).

### 2018-02-09 submission/manuscript.tex
fix study analysis overview

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness as tumor severity increased, after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness translated into small
increases in the odds ratio for individuals to have both adenoma
(P-value = 0.035) and carcinoma tumors (P-value = 0.0021) while the
reduction in Shannon diversity only translated into an increased odds
ratio for individuals to have carcinomas (P-value = 0.0069). Increases
in mouth-associated microbes were commonly in the top 5 most
significantly increased odds ratios for individuals to have either
adenoma or carcinoma tumors regardless of sample type. Prediction models
built to classify either individuals with adenoma or carcinoma were
trained on the whole community or selected genera (top 5 highest and
lowest odds ratios) from either fecal or tissue samples. Both the full
and select models for either adenoma or carcinoma resulted in similar
classification success according to Area Under the Curve (AUC).
Importantly, all models had a generalized decrease in AUC between train
and test data sets. The most important groups within the full community
models consistently belonged to genera such as \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} across studies. Although a
number of associations between the microbiota and tumor were identified,
the majority of studies that we used in this meta-analysis were only
individually adequately powered for large effect sizes.

@@

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity was first assessed for differences between controls,
adenoma stage, and carcinoma stage using both linear mixed-effect models
and RR. The Bray-Curtis index was used to assess, for each individual
study, differences between control-adenoma and control-carcinoma. Next,
all common genera were assessed for differences in RR for having an
adenoma or carcinoma and ranked based on P-value. We then built Random
Forest models based on full or selected community, based on the top 5
increased and top 5 decreased RR based on P-value, and these models were
trained on one study then tested on the remaining studies. This process
was repeated for every study in the meta-analysis. A similar approach
was then applied at the OTU level with the exception that a 10-fold CV
over 100 different models, based on random 80/20 splitting of the data,
was used to generate a range of expected AUCs. For these OTU-based
models, the selected model included all OTUs that had a taxonomic
classification to a taxa in the top 5 increased and top 5 decreased RR
based on P-value. Finally, the power of each study was assessed for an
effect size ranging from 1\% to 30\% and an estimated sample size, for
these effect sizes, was generated based on 80\% power. For comparisons
in which only control versus adenoma stage were made, the carcinoma
samples were excluded from each respective study. Similarly, for
comparisons in which control versus carcinoma stage were made the
adenoma samples were excluded from each respective study. The data were
split between feces and tissue samples. Within the tissue groups the
data were further divided between samples from the same individual
(matched) and those from different individuals (unmatched). Where
applicable for each study, predictions for adenoma and carcinoma stage
of disease were then tested for feces, matched tissue, and unmatched
tissue.

>>>>

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness as tumor severity increased after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness translated into small
increases in the odds ratio for individuals to have both adenoma
(P-value = 0.035) and carcinoma tumors (P-value = 0.0021) while the
reduction in Shannon diversity only translated into an increased odds
ratio for individuals to have carcinomas (P-value = 0.0069). Increases
in mouth-associated microbes were commonly in the top 5 most increased
odds ratios for individuals to have either adenoma or carcinoma tumors
regardless of sample type. Prediction models built to classify either
individuals with adenoma or carcinoma were trained on the whole
community or selected genera (top 5 highest and lowest odds ratios) from
either fecal or tissue samples. Both the full and select models for
either adenoma or carcinoma resulted in similar classification success
according to Area Under the Curve (AUC). The most important groups
within the full community models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Although a number of associations between the microbiota and
tumor were identified, the majority of studies that we used in this
meta-analysis were only individually adequately powered for large effect
sizes.

@@

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity was first assessed for differences between controls,
adenoma stage, and carcinoma stage using both linear mixed-effect models
and OR. For each individual study the Bray-Curtis index was used to
assess differences between control-adenoma and control-carcinoma. Next,
all common genera were assessed for differences in OR for having an
adenoma or carcinoma and ranked based on P-value. We then built Random
Forest models based on the full or selected community (the top 5
increased and top 5 decreased OR based on P-value). Comparison between
the full and selected models took two different approaches. In the first
approach, models were trained on one study then tested on the remaining
studies. This process was repeated for every study in the meta-analysis.
In the second approach models were built using OTU level data and a
10-fold CV over 100 different iterations, based on random 80/20
splitting of the data, was used to generate a range of expected AUCs.
For these OTU-based models, the selected model included all OTUs that
had a taxonomic classification to a taxa in the top 5 increased and top
5 decreased OR based on P-value. Finally, the power of each study was
assessed for an effect size ranging from 1\% to 30\% and an estimated
sample size, for these effect sizes, was generated based on 80\% power.
For comparisons in which only control versus adenoma indivdiuals were
made, the carcinoma individuals were excluded from each respective
study. Similarly, for comparisons in which control versus carcinoma
individuals were made the adenoma individuals were excluded from each
respective study. For all analysis completed feces and tissue samples
were kept seperate. Within the tissue groups the data were further
divided between samples from the same individual (matched) and those
from different individuals (unmatched).
